---
title: "Hierarchical clustering of gene (n=8368) expression clustering for tumours with >=5 samples (n=17)"
author: "Dr Jacob Househam"
date: "18/05/2022"
output: html_document
---

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(DESeq2);library(phangorn);library(wesanderson);library(pheatmap)
library(data.table);library(phylogram);library(dendextend);library(usedist)
options(datatable.fread.datatable=FALSE)
regcol <- c(A='#E31A1C',B='#377DB8',C='#4DAE49',D='#904A9A',Root='#808080');rcol <- regcol[c(1:4)]
```

```{r functions}
getsumdists <- function(dm,sams) {
  regs <- gsub('^C\\d+_(\\S)\\d+_\\S+','\\1',sams)
  dists <- c();for(i in c(1:length(sams))) { for(j in c(1:length(sams))) { if(i<j) {if(regs[i]==regs[j]) { dists <- c(dists,dm[i,j]) } } }}
  return(sum(dists))
}
```

## Load expression data and info
```{r load_expression_and_info}
# 2) Load DESeq2 normalised gene expression
vsd <- readRDS('intermediates/allgenes.vsd.ensembl.rds')
vsd@colData$Region <- gsub('^C\\d+_(\\S)\\S+','\\1',row.names(vsd@colData))
geneexp <- as.data.frame(assay(vsd))

# 3) Input expression filtering from other script
genedf <- readRDS('results/gene_clustering_and_id_conversion.rds')
genedf <- genedf[which(genedf$Group!=4),]

# 4) Load sample data
samples <- read.table('input_files/ListRNAPass.EPICC.txt')[,1]
tsam <- samples[which(gsub('^C\\d+_(\\S)\\d_\\S+$','\\1',samples)!='E')]
patients <- c('C518','C524','C531','C532','C537','C538','C542','C543','C544','C549','C550','C551','C552','C554','C559','C560','C561')
```

## Generate and plot expression clustered dendrograms, plus matching heatmaps
### Plots used in Figures S12 and S14
### To complete Figure S12, take trees from Fig 3 & Figure S28
```{r plot_dna_get_stats}
expdists <- expdends <- list()
pdf('figures/figS12&14.gene_expression_dendrograms.pdf',height=5.556)
par(mar=c(4,4,2,2),font=2,font.axis=2,font.lab=2)
for(pat in patients) {
  vsd_pat <- vsd[which(names(vsd@rowRanges) %in% row.names(genedf)),tsam[grep(pat,tsam)]]
  annodf <- data.frame(Region=vsd_pat@colData$Region);row.names(annodf) <- row.names(vsd_pat@colData)
  distpat <- dist(t(assay(vsd_pat)),method = 'euclidean');hcpat <- hclust(distpat,method='complete')
  expdists[[pat]] <- distpat
  
  patdend <- as.dendrogram(hcpat)
  for(reg in names(rcol)) { patdend <- set(patdend,"by_labels_branches_col", value = row.names(annodf)[which(annodf$Region==reg)], type = "all",TF_value=rcol[reg]) }
  patdend <- set(patdend,"labels_color", value = rcol[annodf$Region],order_value=T)
  patdend <- set(patdend,"branches_lwd", 2)
  labels(patdend) <- gsub('^C\\d+_(\\S+)$','\\1',labels(patdend))
  expdends[[pat]] <- patdend
  # Plotted 'backwards' for Figure S12
  par(mar=c(4,4,2,2),font=2,font.axis=2,font.lab=2)
  plot_horiz.dendrogram(patdend,side = T, main=pat)
  # Plotted right-facing for Figure S14
  par(mar=c(4,2,2,4),font=2,font.axis=2,font.lab=2)
  plot(patdend,horiz = TRUE,main=pat)
  
  # Plot heatmaps to go in Figure S14
  explot <- t(assay(vsd_pat));row.names(explot) <- gsub('^C\\d+_(\\S+_\\S+)$','\\1',labels(patdend))
  sdexp <- apply(explot,2,sd)
  explot <- explot[,which(sdexp!=0)]
  explot <- scale_mat(explot,'column')
  
  mybreaks <- seq(floor(min(explot)),ceiling(max(explot)),by=0.05)
  mat <- pheatmap(explot[rev(labels(patdend)),],show_rownames = T,show_colnames=F,cluster_rows=F,cluster_cols=T,treeheight_col=0,
                  border_color=NA,fontsize = 10,legend = F,scale='none',
                  color=wes_palette("Zissou1", length(mybreaks)-1, type = "continuous"),breaks=mybreaks)
  
}
dev.off()
saveRDS(list(dists=expdists,dends=expdends),file='intermediates/exp_dists_and_dends.rds')
```

## Run permutation test for clustering association with region and plot result
### Figure S15
```{r run_and_plot_permutation}
explist <- readRDS('intermediates/exp_dists_and_dends.rds')

# Reduce to only tumours with at least 10 samples (n=11) - can't get proper distributions with fewer samples
numsams <- c()
for(p in c(1:length(patients))) {
  pat <- patients[p]
  distmat <- as.matrix(explist$dists[[pat]])
  samples <- colnames(distmat);numsams <- c(numsams,length(samples))
}
patients <- patients[which(numsams>10)]

# Run permutation test
resdist <- data.frame(Res=rep(0,length(patients)),Max=0,Prop=0,Sig=0);row.names(resdist) <- patients
permlist <- list()
for(p in c(1:length(patients))) {
  pat <- patients[p]
  distmat <- as.matrix(cophenetic(explist$dends[[pat]]));colnames(distmat) <- row.names(distmat) <- paste0(pat,'_', row.names(distmat))
  samples <- colnames(distmat)
  sumdist <- getsumdists(distmat,samples)
  resdist[pat,'Res'] <- sumdist
  
  permdists <- c()
  for(i in c(1:10000)) {
    tmpmat <- distmat
    colnames(tmpmat) <- row.names(tmpmat) <- sample(samples)
    permdists <- c(permdists,getsumdists(tmpmat,colnames(tmpmat)))
  }
  permlist[[pat]] <- permdists
  
  resdist[pat,'Max'] <- max(permdists);resdist[pat,'Prop'] <- sumdist/max(permdists)
  resdist[pat,'Sig'] <- (length(which(permdists<=resdist[pat,'Res']))+1)/(10000+1)
  
}
resdist$Adj <- p.adjust(resdist$Sig,method='fdr')

# Plot permutation test
pdf(paste0('figures/figS15.mixing_permutations_expression.cophenetic.pdf'),width=6,height=8)
layout(matrix(c(1:12),nrow=4,ncol=3,byrow=T))
for(p in c(1:length(patients))) {
  pat <- patients[p];permdists <- permlist[[pat]]
  
  if(resdist[pat,'Res']<min(permdists)) {
    hist(permdists,breaks=30,xlim=c(resdist[pat,'Res'],max(permdists)),col='skyblue3',border=NA,
         main=pat,xlab=paste0('p=',formatC(resdist[pat,'Adj'])))
  } else {
    hist(permdists,breaks=30,col='skyblue3',border=NA,main=pat,xlab=paste0('p=',formatC(resdist[pat,'Adj'])))
  }
  abline(v=resdist[pat,'Res'],lty=2,lwd=2,col='firebrick3')
}
dev.off()
saveRDS(resdist,file='results/expression_clustering_mixing_results.rds')
```

## Generate expression distances with CIBERSORTx distances
```{r generate_distances}
ciber <- read.delim("input_files/CIBERSORTx_EPICC.txt");ciber <- ciber[ciber$Mixture %in% tsam,]
row.names(ciber) <- ciber$Mixture
celltypes <- colnames(ciber[,c(2:23)])
ciber$Patient <- as.factor(gsub("^(C\\d+)_\\S+_\\S+","\\1",ciber$Mixture))

# Generate immune Euclidean distance matrices
immdists <- list()
for(pat in patients) {
  ciberpat <- ciber[tsam[grep(pat,tsam)],celltypes]
  
  distpat <- dist(ciberpat,method = 'euclidean')
  immdists[[pat]] <- distpat
}

# Get all pairwise distances for expression and immune
disexp_dists <- data.frame(matrix(ncol=7,nrow=0));colnames(disexp_dists) <- c('s1','s2','patient','pair','pair_type','exp_d','ciber_d')
for(pat in patients) {
  curpat <- tsam[grep(pat,tsam)]
  for(s1 in c(1:length(curpat))) {
    sam1 <- curpat[s1]
    for(s2 in c(1:length(curpat))) {
      if(s1<s2) {
        sam2 <- curpat[s2]
        curexpd <- as.numeric(dist_subset(expdists[[pat]], c(sam1,sam2)))
        reg1 <- gsub('C\\d+_(\\S)\\S+','\\1',sam1);reg2 <- paste0(gsub('C\\d+_(\\S)\\S+','\\1',sam2))
        type <- ifelse(reg1==reg2,'within-regions','between-regions')
        
        curcib <- as.numeric(dist_subset(immdists[[pat]], c(sam1,sam2)))
        
        disexp_dists[nrow(disexp_dists)+1,] <- c(sam1,sam2,pat,paste0(reg1,'-',reg2),type,curexpd,curcib)
      }
    }
  }
}
disexp_dists$exp_d <- as.numeric(disexp_dists$exp_d)
disexp_dists$ciber_d <- as.numeric(disexp_dists$ciber_d)
```

## Plot comparison of expression and immune based distances
### Figure S16 (A-C)
```{r plot_compare_distances}
datacol <- brewer.pal(3,'Accent')[c(1:2)]
res <- lm(disexp_dists$ciber_d~disexp_dists$exp_d);sumres <- summary(res)

pdf('figures/figS16A.CIBER_vs_Expression_Distance.pdf')
par(las=1,cex.axis=1.5,cex.lab=1.5,font=2,font.lab=2,font.axis=2,mar=c(4.5,5,2.5,1))
plot(disexp_dists$exp_d,disexp_dists$ciber_d,pch=16,
     col=scales::alpha(ifelse(disexp_dists$pair_type=='within-regions',datacol[1],datacol[2]),0.6),xlim=c(0,500),ylim=c(0,30),bty='n',xlab='Expressed Genes',ylab='CIBERSORT Absolute Score')
abline(a=as.numeric(res$coefficients[1]),b=as.numeric(res$coefficients[2]),lty=2,lwd=3);mtext(side=3,text='Intra-tumour euclidean distances',cex=1.5)
legend('topright',bty='n',
       legend=c(paste0('p=',formatC(sumres$coefficients[2,4], format = "e", digits = 2)),paste0('R^2=',signif(sumres$r.squared,3))))
legend('topleft',bty='n',col=datacol,pch=16,
       legend=c('within-regions','between-regions'),cex=1.35)
dev.off()

disexp_dists$pair_type <- factor(disexp_dists$pair_type,levels=c('within-regions','between-regions'))
pdf('figures/figS16BC.CIBER_and_Expression_violins.pdf',width=5.5,height=5.5)
res <- wilcox.test(exp_d~pair_type,data=disexp_dists)
datacol <- brewer.pal(3,'Accent')[c(1:2)]
par(mar=c(5,5,2,1),font.axis=2,font=2,font.lab=2,cex.lab=1.5,cex.axis=1.2)
vioplot(exp_d~pair_type,data=disexp_dists,col=datacol,xlab='Comparison type',ylab='Expressed Genes Euclidean Distance',
        ylim=c(0,600),border=datacol,axes=F,las=1,drawRect=T)
grid(lty=1)
vioplot(exp_d~pair_type,data=disexp_dists,col=datacol,ylim=c(0,1),border=datacol,axes=F,las=1,drawRect=T,add=T)
mtext(side=3,text=paste0('wilcoxon p-value: ',formatC(res$p.value, format = "e", digits = 2)),cex=1.2)

res <- wilcox.test(ciber_d~pair_type,data=disexp_dists)
vioplot(ciber_d~pair_type,data=disexp_dists,col=datacol,xlab='Comparison type',ylab='CIBERSORTx Euclidean Distance',
        ylim=c(0,30),border=datacol,axes=F,las=1,drawRect=T)
grid(lty=1)
vioplot(ciber_d~pair_type,data=disexp_dists,col=datacol,ylim=c(0,1),border=datacol,axes=F,las=1,drawRect=T,add=T)
mtext(side=3,text=paste0('wilcoxon p-value: ',formatC(res$p.value, format = "e", digits = 2)),cex=1.2)
dev.off()
```

