---
title: "Run and plot gene expression phylogenetic signal analysis"
author: "Dr Jacob Househam"
date: "21/06/2021"
output: html_document
---

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(phytools);library(DESeq2);library(org.Hs.eg.db);library(phangorn);library(wesanderson);library(pheatmap);library(STRINGdb);library(metap);library(clusterProfiler);library(stringr)
```

```{r setup}
ngroupcol= c(wes_palette("Darjeeling2",5)[2],wes_palette('Darjeeling1',5)[2],wes_palette('FantasticFox1',5)[5])
names(ngroupcol) <- as.numeric(c(1:3))
col_list <- list(Group=ngroupcol)
string_db <- STRINGdb$new(version="11",species=9606,score_threshold=200,
                          input_directory="~/Documents/EPICC/PurityAdjust/stringdb_dir/")
                          #input_directory="<path_to_downloaded_string_db_database>")
```

## Load tree and expression data
```{r load_tree_and_expression}
# 1) Load tree info
treelist <- readRDS('intermediates/pgls.trees.rds')
tinfo <- readRDS('intermediates/pgls.infotrees.rds')

# 2) Load DESeq2 normalised gene expression
vsd <- readRDS('intermediates/allgenes.vsd.ensembl.rds')
geneexp <- as.data.frame(assay(vsd))[,tinfo$samples]

# 3) Input expression filtering from other script
genedf <- readRDS('results/gene_clustering_and_id_conversion.rds')
```

## For each tumour get the phylogenetic signal (lambda) for all genes for the 100 trees of varying branch length
### Takes a long time - should be run on an HPC
```{r get_phylogenetic_signal}
gene_lambda_list <- list()
for(pat in tinfo$patients) {
  trimtree <- treelist[[pat]][[1]]
  treesam <- paste0(pat,"_",trimtree$tip.label)
  print(paste0(pat,': ',length(treesam),' matched samples x ',nrow(genedf),' genes'))
  
  try <- as.data.frame(t(geneexp[row.names(genedf),treesam]))
  row.names(try) <- gsub('C\\d+_(\\S+)','\\1',row.names(try))
  
  siglist <- list()
  for(i in c(1:length(treelist[[pat]]))) {
    trimtree <- treelist[[pat]][[as.character(i)]]
    
    sigphylo <- data.frame(Gene=colnames(try),Lambda=0,Lpval=0,Kstat=0,Kpval=0)
    for(j in c(1:ncol(try))) {
      curgen <- try[,j];names(curgen) <- row.names(try)
      if(var(curgen)>0) {
        res <- phylosig(trimtree,curgen,method='lambda',test=T)
        sigphylo[j,'Lambda'] <- res$lambda;sigphylo[j,'Lpval'] <- res$P
      } else {
        res <- phylosig(trimtree,curgen,method='lambda',test=F)
        sigphylo[j,'Lambda'] <- res$lambda;sigphylo[j,'Lpval'] <- 1
      }
    }
    siglist[[as.character(i)]] <- sigphylo
    print(paste0(pat,': ',signif((i/length(treelist[[pat]])*100),2),'%'))
  }
  
  lam <- do.call(cbind, lapply(siglist,function(x) { x$Lambda }))
  lam_pval <- do.call(cbind, lapply(siglist,function(x) { x$Lpval }))
  resdf <- data.frame(MedLambda=rowMedians(lam),
                    LamPval=rowMedians(lam_pval))
  row.names(resdf) <- colnames(try)
  
  gene_lambda_list[[pat]] <- resdf
}
saveRDS(gene_lambda_list,file='results/gene_lambda_data.rds')
```

## Analyse and plot the results of the phylogenetic gene expression analysis
### Figure 1G&H and Figure S3)
```{r plot_phylogenetic_gene_results}
genedf <- genedf[which(genedf$Group!='4'),]
reclam <- matrix(0L,nrow=nrow(genedf),ncol=length(tinfo$patients))
colnames(reclam) <- tinfo$patients;row.names(reclam) <- row.names(genedf)
recstars <- recpval <- recadjstar <- reclam
pdf('figures/fig1GH.lambda_analysis_genes.pdf',height=8,width=5)
for(pat in tinfo$patients) {
  resdf <- gene_lambda_list[[pat]]
  resdf <- resdf[row.names(genedf),]
  
  combdf <- merge(resdf,genedf,by=0);row.names(combdf) <- combdf$Row.names
  combdf$MedPval <- -log10(combdf$LamPval)
  reclam[row.names(combdf),pat] <- combdf$MedLambda
  recstars[row.names(combdf),pat] <- getstars_vec(combdf$LamPval)
  recpval[row.names(combdf),pat] <- combdf$MedPval
  recadjstar[row.names(combdf),pat] <- getstars_vec(p.adjust(combdf$LamPval,method='fdr'))
  
  trimtree <- treelist[[pat]][[1]]
  treesam <- paste0(pat,'_',trimtree$tip.label)
  try <- as.data.frame(t(geneexp[row.names(resdf),treesam]))
  row.names(try) <- gsub('C\\d+_(\\S+)','\\1',row.names(try))
  
  seldf <- combdf[which(combdf$LamPval<0.05),]
  
  regions <- gsub('^(\\S)\\d+_\\S+','\\1',trimtree$tip.label)
  options(scipen = -1);par(mar=c(0,0,2,0),xpd=T)
  plot.phylo(trimtree,type="phylogram",align.tip.label = T,edge.width=3,
             font=2,cex=0.9,tip.color=regcol[c(regions,'Root')],label.offset=20)
  title(main=pat)
  
  if(nrow(seldf)>1) {
    seldf <- seldf[seldf$hgnc_symbol!='',]
    explot <- try[,row.names(seldf)];colnames(explot) <- seldf$hgnc_symbol
    explot <- scale_mat(explot,'column')
    d = dist(t(explot),method='euclidean');clustord <- hclust(d,method='complete')
    explot <- explot[c(nrow(explot):1),clustord$order]
    
    newcolannot <- seldf;row.names(newcolannot) <- newcolannot$hgnc_symbol
    mybreaks <- seq(floor(min(explot)),ceiling(max(explot)),by=0.05)
    pheatmap(explot,show_rownames = F,show_colnames=F,cluster_rows=F,cluster_cols=F,treeheight_col=0,
             border_color=NA,fontsize = 6,fontsize_col = 4,
             color=wes_palette("Zissou1", length(mybreaks)-1, type = "continuous"),breaks=mybreaks,
             annotation_col=newcolannot[,c('MedLambda','MedPval','Group')],
             annotation_colors=col_list)
  }
}
dev.off()

recsave <- genedf;recsave$NumRec <- rowSums(recstars!='')
saveRDS(recsave,file='results/reccurent_phylogenetic_genes.rds')
```

## Plot recurrent phylogenetic genes
### Figure 1I
```{r plot_rec_phylo}
geneanot <- recsave[which(rowSums(recstars!='')>=3),];geneanot <- geneanot[order(geneanot$NumRec,decreasing=T),]
filrecstars <- recstars[row.names(geneanot),];row.names(filrecstars) <- geneanot$hgnc_symbol
filrecpval <- recpval[row.names(geneanot),];row.names(filrecpval) <- geneanot$hgnc_symbol
row.names(geneanot) <- geneanot$hgnc_symbol
geneanot <- geneanot[,c("Mean_Mean_Exp","Group","NumRec")]

pdf('figures/fig1I.recurrent_genes_lambda.pdf',height=12,width=8.5)
mybreaks <- seq(0,ceiling(max(filrecpval)),by=0.01)
pheatmap(filrecpval,show_rownames=T,show_colnames=T,cluster_rows=F,cluster_cols=T,
         border_color=NA,fontsize=15,fontsize_col=20,fontsize_row=11,
         color=wes_palette("Zissou1", length(mybreaks)-1, type = "continuous"),breaks=mybreaks,
         display_numbers=filrecstars,fontsize_number=15,
         annotation_row=geneanot,
         annotation_colors=col_list)
dev.off()
```

## Plot chi-squared test of phylo vs. gene groups
### Figure 1J
```{r plot_chisquared_test}
recsave$Sig <- ifelse(recsave$NumRec>=1,'Sig','NotSig')
pdf('figures/fig1J.gene_phylo_chiplot.pdf')
par(mar=c(4.5,4.5,2,2),font=2,font.axis=2,font.lab=2,cex.axis=2)

mat <- matrix(0L,nrow=3,ncol=2,dimnames=list(c('3','2','1'),c('Sig','NotSig')))
probmat <- mat
for(cat in as.character(c(1:3))) {
  mat[cat,'NotSig'] <- length(which(recsave$Group==cat & recsave$Sig=='NotSig'))
  mat[cat,'Sig'] <- length(which(recsave$Group==cat & recsave$Sig=='Sig'))
  
  probmat[cat,] <- mat[cat,]/sum(mat[cat,])*100
}
res <- chisq.test(mat)  

xx <- barplot(t(probmat),beside=T,xlim=c(0,100),border=c(ngroupcol[3],ngroupcol[3],ngroupcol[2],ngroupcol[2],ngroupcol[1],ngroupcol[1]),axes=F,horiz=T,las=2,density=c(NA,20,NA,20,NA,20),
              col=c(ngroupcol[3],ngroupcol[3],ngroupcol[2],ngroupcol[2],ngroupcol[1],ngroupcol[1]))
axis(side=1,cex.axis=1.75,line=0);mtext(side=2,text='Group',line=2.75,cex=1.6)
mtext(side=1,text='Percentage of genes',line=2.75,cex=1.6)
mtext(side=3,text=paste0('chisq test: X^2=',signif(res$statistic,3),' df=',res$parameter,', p-value=',signif(res$p.value,3)),line=0.5,cex=1.2)
legend(x=70,y=3.5,xpd=T,cex=1.5,legend=c('NonPhylo','Phylo'),bty='n',fill='gray60',density=c(20,NA),border=c('gray60'))
dev.off()
```

## Plot stringDB analysis of recurrently phylogenetic genes
### Figure 1K
```{r plot_chisquared_test}
stringdf <- string_db$map(recsave, "ensembl_gene_id", removeUnmappedRows = TRUE)
enrich <- string_db$get_enrichment(stringdf$STRING_id[which(stringdf$NumRec>=3)],category='KEGG')
# Write string IDs of significant PPAR signaling pathway to upload to string-db.org in order
# to make Figure 3K
write.table(str_split(enrich[which(enrich$description=='PPAR signaling pathway'),'inputGenes'],',')[[1]],file='results/stringdb.rec.ppar.txt',row.names = F,col.names=F,quote=F,sep='\n')
```

## Plot the (non-)correlation of sample number x number of sig phylo gene
### Figure S4
```{r plot_correlation}
res <- lm(colSums(recstars!='')~tinfo$numsam);sumres <- summary(res)
pdf('figures/figS4.correlation_plot_number_samples_vs_sig_phylo_genes.pdf')
par(mar=c(4,4,1,1),font=2,font.axis=2,font.lab=2)
plot(tinfo$numsam,colSums(recstars!=''),bty='n',pch=16,cex=2,col='skyblue3',
     xlab='',ylab='',axes=F,xlim=c(0,35),ylim=c(0,2500))
axis(side=1);axis(side=2)
mtext(side=1,text='Number of tumour samples',line=2.5)
mtext(side=2,text='Number of significant phylogenetic genes',line=2.5)
text(tinfo$numsam,colSums(recstars!=''),labels=tinfo$patients,pos=c(4,3,3,3,4,3,2,1),offset=0.6,cex=1)
abline(a=res$coefficients[1],b=res$coefficients[2],lty=2,col=scales::alpha('black',0.75))
legend(x=26,y=1600,legend=c(paste0('adj-R^2 = ',signif(sumres$adj.r.squared,2)),paste0('p = ',signif(sumres$coefficients[2,4],2))),
       bty='n',col=scales::alpha('black',0.75),cex=0.9)
dev.off()
```


